---
title: "Gun Laws"
author: "Anam Khan"
date: "10/22/2021"
output:
  pdf_document: default
  html_document: default
---
# This file contains general EDA and code to support Research Question #3

### Reading in gun law by state Data
```{r}
library(ggplot2)
library(dplyr)
library(randomcoloR)

###############################################
##     STATE FIREARM LAWS NATIONAL DATA      ##
## http://www.statefirearmlaws.org/resources ##
###############################################

# Read Dataset in 
gunLaws <- read.csv("gunLawsBU.csv", header = TRUE, na.strings = c("", " ", "N/A"))
# head(gunLaws)

# Look at data types
str(gunLaws)
states <- levels(gunLaws$state)

# For each row, get the total of all gun laws for that state for that year
for (i in 1:nrow(gunLaws)){
  gunLaws$total[i] <- sum(gunLaws[i, 3:137])
}

# Subset the DF to only have the State, Year, and Totals
wantCols <- c(1,2,138)
# gunLaws[,wantCols]

# Split the OG dataset into 50 smaller data sets (years and totals by ST)
splitBySt <- split(gunLaws, gunLaws$state)
length(splitBySt)
 
# random list of 50 colors 
myColors <- randomColor(50)

# Getting total number of laws for a random state (MD)
MD <- as.data.frame(splitBySt["Maryland"])
MD[MD$Maryland.year == "2018", "Maryland.total"]

# loop through each state and plot its laws
# for (i in 1:length(splitBySt)){
#   new <- data.frame(splitBySt[i])
#   new <- new[,wantCols]
#   plot(new[,2], new[,3],
#       type = "l", pch = 19, lwd =2, col = myColors[i],
#       main = new[1,1], xlab = "Year", ylab = "Number of Laws/Provisions")
# 
# }

```

# Mapping Gun Violence Incidents 
```{r}
##############################################
##      US MASS SHOOTINGS 1982 - 2021       ##
## https://data.world/thebleak/thebleak13s1 ##
##############################################

# Read Dataset in 
massShootings <- read.csv("Mother Jones US Mass Shootings 1982_2021.csv", header = TRUE, na.strings = c("", " ", "N/A"))
head(massShootings)
table(massShootings$state)

# Split OG DF into dataframes by state
massShootings_split <- split(massShootings, massShootings$state)
# 37 states are accounted for here
length(massShootings_split)
# head(massShootings_split)

# Get the number of mass shootings per state
numShootins_State <- sapply(massShootings_split, nrow)
# Call number mass shooting for any random state 
numShootins_State["Florida"]

library(ggmap)
# Map of all mass shootings
qmplot(longitude, latitude, data = massShootings, colour = I('red')) 

# All mass shooting without Hawaii (1 data point)
massShootings_xHawaii <- massShootings[-95,]
qmplot(longitude, latitude, data = massShootings_xHawaii, colour = I('red')) 

# California
# caliDF
# typeof(caliDF)
# qmplot(California.longitude, California.latitude, data = caliDF , colour = I('red'),  size = I(2))

# Florida 
floridaDF <- as.data.frame(massShootings_split["Florida"])

qmplot(Florida.longitude, Florida.latitude, data = floridaDF , colour = I('red'),  size = I(2))
```

## Gun Violence and partisan control of the Senate
Does the political party controlling the Senate have any impact on number of individuals killed in that state? 
```{r}
#######################################################
##             GUN VIOLENCE 2013 - 2018              ##
## https://www.kaggle.com/jameslko/gun-violence-data ##
#######################################################

gunViolence_kag <- read.csv("gun-violence-data_01-2013_03-2018.csv", header = T, na.strings = c("", " ", "N/A"))

head(gunViolence_kag)


keepCols <- c(2:4, 6:7)
StategunViolence_kag <- gunViolence_kag[, keepCols]
head(StategunViolence_kag)

gunViolence_numKilled.State <- aggregate(n_killed ~ state, data = StategunViolence_kag, FUN = sum)

gunViolence_numKilled.State <- as.data.frame(gunViolence_numKilled.State)
gunViolence_numKilled.State <- gunViolence_numKilled.State[order(-gunViolence_numKilled.State[,"n_killed"]),]

library(RColorBrewer)

## Red or Blue States (partisan control of state senates) - senate laws
## https://ballotpedia.org/Partisan_composition_of_state_legislatures

BlueStates <- c('California',	'Delaware',	'Hawaii',	'New Jersey',	'Oregon',	'Rhode Island',	'Washington')

RedStates <- c('Alabama',	'Alaska',	'Arizona',	'Arkansas',	'Florida',	'Georgia',	'Idaho',	'Indiana',	'Iowa',	'Kansas',	'Kentucky',	'Michigan',		'Mississippi',	'Missouri',	'Nebraska',	'New Hampshire',	'North Dakota',	'Ohio',	'Oklahoma',	'South Carolina',	'South Dakota',	'Tennessee',	'Texas',	'Utah',	'West Virginia',	'Wisconsin',	'Wyoming')

PurpleStates <- c('Colorado', 'Connecticut', 'Illinois', 'Maine',	'Maryland', 'Louisiana', 'Massachusetts', 'Minnesota', 'Montana' ,'Nevada',	'New Mexico',	'New York',	'North Carolina','Pennsylvania',	'Vermont',	'Virginia')

AllStates <- c('Alabama',	'Alaska',	'Arizona',	'Arkansas',	'Florida',	'Georgia',	'Hawaii',	'Idaho',	'Indiana',	'Iowa',	'Kansas',	'Kentucky',	'Louisiana',	'Michigan',	'Minnesota',	'Mississippi',	'Missouri',	'Montana'	,'Nebraska',	'New Hampshire',	'North Carolina',	'North Dakota',	'Ohio',	'Oklahoma'	,'Pennsylvania',	'South Carolina',	'South Dakota',	'Tennessee',	'Texas',	'Utah',	'West Virginia',	'Wisconsin',	'Wyoming', 'California',	'Colorado',	'Connecticut',	'Delaware',	'Illinois',	'Maine',	'Maryland',	'Massachusetts',	'Nevada',	'New Jersey',	'New Mexico',	'New York',	'Oregon',	'Rhode Island',	'Vermont',	'Virginia',	'Washington')

StateParty <- data.frame(AllStates)

gunViolence_numKilled.State[1] <- lapply(gunViolence_numKilled.State[1], as.character)
# gunViolence_numKilled.State[3] <- lapply(gunViolence_numKilled.State[3], as.character)

for ( i in 1:nrow(gunViolence_numKilled.State)){
  if (gunViolence_numKilled.State$state[i] %in% BlueStates){
    gunViolence_numKilled.State$polParty[i] <- "Blue"
  }
  else if (gunViolence_numKilled.State$state[i] %in% RedStates){
    gunViolence_numKilled.State$polParty[i] <- "Red"
  }
  else if (gunViolence_numKilled.State$state[i] %in% PurpleStates){
    gunViolence_numKilled.State$polParty[i] <- "Purple"
  }
}

gunViolence_numKilled.State[3] <- lapply(gunViolence_numKilled.State[3], as.factor)
gunViolence_numKilled.State <- as.data.frame(gunViolence_numKilled.State)

gunViolence_numKilled.State <- gunViolence_numKilled.State[gunViolence_numKilled.State$state != 	"District of Columbia",]


######################################################
# Total Number killed '13-'18 by State (& pol party) #
######################################################

ggplot(data = gunViolence_numKilled.State, 
       aes(x = reorder(state, n_killed), y=n_killed), colour = 
       as.factor(gunViolence_numKilled.State$polParty)) + 
  geom_bar(stat = "identity", 
           aes(fill = polParty)) + 
  coord_flip() +
  theme_minimal() +
  ggtitle("Individuals Killed due to Gun Violence per State (2013-2018)") +
  xlab("Total Number Killed") +
  ylab("State") +
  scale_x_discrete(limits = rev(levels(gunViolence_numKilled.State$state))) +
  geom_text(aes(label=n_killed), 
            position=position_dodge(width=0.9),
            hjust=-0.25, size = 2.5) +
  scale_fill_manual(values = c("blue", "purple", "red"), labels = c("Democratic", "Divided", "Republican")) +
  guides(fill=guide_legend(title="Political Party \n(partisan control of Senate)")) +
  theme(legend.title = element_text(size = 8),
        legend.background = element_rect(fill = "white"),
        legend.position = c(0.85, 0.15))
  
```

## Hypothesis Testing

We will use ANOVA to analyze since our variable of interest "political party" has more than 2 levels. If we just had Dem and Rep, then we would do t-test

```{r}
######################
# Hypothesis Testing #
######################

#################
# 1 - WAY ANOVA #
#################

gunViolence_numKilled.State


# one way table
table(gunViolence_numKilled.State$polParty)

# saving the same table into a new variable for anova 
anova_GV_nk_state <- gunViolence_numKilled.State[-1]

# Visualizing differences with side by side boxplots
plot(n_killed ~ polParty, data = anova_GV_nk_state, 
     col = c("steelblue1", "mediumorchid1", "red2"),
     xlab = "Partisan Control of the Senate", 
     ylab = "Total number killed",
     names = c("Democratic" ,"Divided", "Republican"))

# lets try to normalize 
#define Min-Max normalization function
min_max_norm <- function(x) {
    (x - min(x)) / (max(x) - min(x))
  }

anova_GV_nk_state_norm <- anova_GV_nk_state
anova_GV_nk_state_norm[1] <- as.data.frame(lapply(anova_GV_nk_state_norm[1], min_max_norm))

head(anova_GV_nk_state_norm)

# ANOVA
polParty.aov <- aov(n_killed ~ polParty, data = anova_GV_nk_state_norm)
summary(polParty.aov)
    # high p-value suggests that all of the variances are the same and we fail to reject our null 


library(ggpubr)
# look at normalization of the data 
#   normal data is a stipulation of the anova test
qqb4 <- ggqqplot(gunViolence_numKilled.State$n_killed, main = "Total killed non-normalized") 
qqafter <- ggqqplot(anova_GV_nk_state_norm$n_killed, main = "Total killed normalized")

densb4 <- ggdensity(gunViolence_numKilled.State$n_killed, main = "Total killed non-normalized")
densafter <- ggdensity(anova_GV_nk_state_norm$n_killed, main = "Total killed normalized")

ggarrange(qqb4, qqafter, densb4, densafter, ncol = 2, nrow = 2)




plot(n_killed ~ polParty, data = anova_GV_nk_state_norm, 
     col = c("steelblue1", "mediumorchid1", "red2"),
     xlab = "Partisan Control of the Senate", 
     ylab = "Total number killed",
     names = c("Democratic" ,"Divided", "Republican"))

polParty.aov.norm <- aov(n_killed ~ polParty, data = anova_GV_nk_state_norm)
summary(polParty.aov.norm)

```


When anova normality conditions are not met fully
```{r}
##################
# KRUSKAL WALLIS #
##################

anova_GV_nk_state

# Mean plots
ggline(anova_GV_nk_state, x = "polParty", y = "n_killed", 
       add = c("mean_se", "jitter"), 
       order = c("Blue", "Purple", "Red"),
       ylab = "Total number killed", xlab = "Political ")

# kruskal test
kruskal.test(n_killed ~ polParty, data = anova_GV_nk_state)

```

p-value is greater than significance level of 0.05 we can conclude that there is not a significant differences between the political party.


Kruskal-Wallis test by rank is a non-parametric alternative to one-way ANOVA test, which extends the two-samples Wilcoxon test in the situation where there are more than two groups. It’s recommended when the assumptions of one-way ANOVA test are not met. http://www.sthda.com/english/wiki/kruskal-wallis-test-in-r


## Wilcoxon test 

The unpaired two-samples Wilcoxon test (also known as Wilcoxon rank sum test or Mann-Whitney test) is a non-parametric alternative to the unpaired two-samples t-test, which can be used to compare two independent groups of samples. It’s used when your data are not normally distributed.
http://www.sthda.com/english/wiki/unpaired-two-samples-wilcoxon-test-in-r
```{r}
# Copy dataframe
t_GV_nk_state <- gunViolence_numKilled.State

# get rid of divided states (purple states)
t_GV_nk_state <- t_GV_nk_state[t_GV_nk_state$polParty != "Purple",]
t_GV_nk_state <- t_GV_nk_state[-1]

wilcox.test(n_killed ~ polParty , data = t_GV_nk_state, exact = F)
wilcox.test(n_killed ~ polParty , data = t_GV_nk_state, exact = F, alternative = "less")
```


High p-value again 





